---
title: "Analysis of COVID-19 phases and virulence 2020-2024"
date: today
date-format: "DD MMMM, YYYY"
lang: en-US
thanks: Thank you, BST 260 teaching team, for your hard work throughout the semester. 
author: 
  - Charles Bray\thanks{SM Candidate, Department of Biostatistics, Harvard TH Chan School of Public Health, \href{mailto:bray@hcp.med.harvard.edu}{\texttt{bray@hcp.med.harvard.edu}}}
abstract: |
  \noindent This study aimed to assess the evolution of the SARS-CoV 2 virus, its associated disease burden, and the public health response. Data from the CDC on hospitalizations, deaths, and cases were analyzed and divided into phases based on these metrics, and a novel approach using a sliding window of 9 weeks was developed to define pandemic waves. Seven waves of the pandemic were identified from January 2020 to December 2024, over which case, hospitalization, and death rates varied across states and regions. The virulence of COVID-19 decreased over time, with the mortality-to-case ratio dropping from 0.06 in the first wave to 0.01 in later waves while the mortality-to-hospitalization ratio dropped from 0.5 to 0.1. This study highlights the importance of standardized definitions for pandemic phases and provides insights into regional public health responses and their consequences. These definitions and associated state-level performance measures provide valuable historical context for future pandemic readiness.
abstractspacing: double
keywords: R, \LaTeX, Quarto
appendix: false
fontfamily: libertine
monofont: inconsolata
monofontoptions: scaled=.95
fontsize: 12pt
spacing: double
geometry: 
  - top=2cm
  - bottom=2cm
  - left=2cm
  - right=2cm
urlcolor: darkblue
linkcolor: Mahogany
citecolor: Mahogany
highlight-style: arrow
pdf-engine: pdflatex
format: 
    pdf:
      template: article-template.latex
      toc: false
      toc-depth: 3
      number-sections: true
      css: styles.css
      cite-method: biblatex
      bibliography: references.bib
editor:
  render-on-save: true
  markdown: 
    wrap: 72
---

```{r, warning=FALSE, include=FALSE}

### prep libraries and data

library(magrittr)
library(dplyr)
library(data.table)
library(kableExtra)

casewavedef <- fread("../figs/Table1.csv") %>% select(begindate,enddate) %>% unique()

minwavebegin <- as.character(min(pull(casewavedef,begindate)))
minwaveend <- min(pull(casewavedef,enddate))
wavecount <- n_distinct(pull(casewavedef,begindate))

statedeathtab <-
  fread("../figs/Table2.csv") %>%
  filter(StateRank<=10) %>%
  .[,c(1:8)] %>% rename(`Mortality (rank)`=StateRank)

names(statedeathtab) <- gsub("state_","Wave starting ",names(statedeathtab))

virulence <- fread("../figs/Table3.csv")

```

\newpage

# Introduction {#sec-introduction}

The COVID-19 pandemic, in which the world saw the emergence of an acute
upper respiratory infection caused by the novel coronavirus SARS-CoV-2,
has been a major cause of death and disability in the United States and
globally since the first months of 2020. Over 1 million deaths are
reported through time of writing in the United States, according to the
World Health Organization, while millions are suspected to have
diminished quality of life from the still poorly-understood set of
conditions associated with "long COVID". (@WHOcite) (@longCovid). It is
popular to describe the net consequences of the pandemic but less so to
assess the evolution of the virus and the evolving ability of our public
health and medical sectors to respond to the morbidity and mortality it
caused, as measured through deaths, hospitalizations, and rates of
infection.

Systematizing the way in which we assess individual phases of a pandemic
is important for several reasons: \newline

*Standardized definitions*: Researchers can compare data and findings in
a consistent way rather than contrasting conclusions that rely on novel
estimations of pandemic phases.

*Inter-regional comparability*: With consistent wave definitions, it is
possible to compare regional or state-level responses within the U.S.
This allows for the identifcation of what works well and what needs to
improve as far as local public health responses (e.g., were school
lock-downs too long or too short in certain areas, what was the effect
of masking during each phase of the pandemic, etc.).

*Clear historical record of pandemic trajectory*: By clearly defining
waves, a framework is created for understanding how the pandemic changed
over time, which allows us to contextualize the public health response
based on how the virus's behavior and human responses evolved.

*Future pandemic readiness*: The identification of distinct waves gives
public health officials and other relevant decisionmakers historical
context that leads to better planning and resource allocation in
anticipation of future COVID-19 surges or other viruses that may behave
similarly. \newline

Some investigators have aimed to address this gap in knowledge by
defining "waves", or distinct phases characterizing the COVID-19
pandemic by the pattern or trajectory of infection as well as the
predominant viral strain. For instance, Iran's experience of the
pandemic was characterized by five distinct phases according to Amin et
al. (2022) through the end of 2022, as defined by public health
communications and increases in case counts following times of
stability. (@Iranwaves). Another group of researchers in Chile defined
the start and end of waves based on sustained increase or decrease as
well as case rates achieving a certain threshold. (@Chilewaves). Other
researchers employed more sophisticated methods, including Break Least
Squares, to define two waves in 2020 in North America each lasting
roughly 50 days. (@covidtimecite). However, little research exists to
clarify times of transition between waves of COVID-19 in the United
States, strictly speaking.

To this end, it is necessary to provide a conceptualization of the
distinct stages of the COVID-19 pandemic as it affected the US
population, as well as the relative performance of different areas
within the US. This paper aims to describe the changing dynamics of the
novel coronavirus as such.

# Methods {#sec-methods}

Data were obtained from the Center for Disease Control and Prevention
(CDC) via API, comprising hospitalizations, deaths, and cases associated
with COVID-19 infection and reported weekly at the US state-level.
(@CDCdata). States were defined as the 50 standard US states along with
DC and Puerto Rico, for a total of 52 geographical units; states were
assigned to regions of the US based on a classification scheme provided
on the BST 260 course GitHub repository. (@rafalab).

The study proceeds in three parts. First, the COVID-19 pandemic was
split broadly into different phases based on counts of cases, deaths,
and hospitalizations across regions of the United States. Following
this, the performance of individual US states was described along these
measures within each determined wave. Finally, the nature of COVID-19
strains (their virulence and/or strain on hospitals) is determined by
comparing the evolution of different measures from early to later waves.

In order to describe the trajectory of the COVID-19 pandemic, a novel
approach to defining breakpoints between infection regimes (or "waves"
of COVID-19) is developed. A sliding window of 9 weeks (or roughly 2
months) was applied to each measure (hospitalizations, cases, and
deaths) within region, in which window were determined if the center
value (the 5th index) of the window was a maximum or minimum value.
Then, if at least half of the regions reporting data (typically, 10
regions) experienced a local minimum within 5 weeks of each other (or,
half the minimum/maximum search window), the current wave was determined
to have ended. Instead of over-weighting high-population regions, this
approach appreciates potential regional segmentation of the pandemic
when evaluating country-level phases of viral spread.

Prior to algorithm development, (1) break least squares as well as (2)
Markov Regime Switching (MRS) models were considered (@covidtimecite).
Ultimately, the complexity of these models–and in the case of MRS, the
infeasibility of constraining wave definitions to just two
regimes–motivated the development of this data-driven identification
strategy. Other, more simple, strategies such as that featured in Ayala
et al., 2021 (@Chilewaves), were also considered but determined to be
too inflexible. Namely, hard thresholds for number of cases per
population were used in defining transitions in infectious regimes,
which were seen here as not considerate of changes in testing ubiquity
and changing virulence of different generations of COVID-19.

# Results {#sec-results}

**Defining waves**

The COVID-19 pandemic in the US was divided into 7 waves from January 1,
2020 through December 1, 2024 according to case and hospitalization
rates. The choice of these two measures as opposed to deaths was
determined to be appropriate based on the persistent lag of deaths data
relative to cases and hospitalizations, and the fact that death timing
relative to time of infection is known to exhibit much greater variance
than symptoms or hospitalization timing relative to time of infection.
(@deathtiming). See Figure 1 below for a visualization of wave
boundaries, where solid black dots indicate local minima in
per-population rates of cases or hospitalizations by region, while black
crosses indicate local maxima. Each dotted red line shows the
inter-regional wave cutoff (where one wave ends and another begins);
there is no assumed interim period between waves (i.e., one wave ends
the day before another begins).

![Definition of candidate COVID-19 waves based on regional minima and
maxima](%22../figs/Figure1.png%22){fig-align="center"}

The first wave occurred roughly March 2020 to June 6, 2020, which we
classify as the "Originator" wave. This was succeeded by the Second and
Third waves, which occurred `r casewavedef$begindate[2]` to
`r as.Date(casewavedef$enddate[2])` and `r casewavedef$begindate[3]` to
`r as.Date(casewavedef$enddate[3])`, respectively. The fourth wave,
which we can call the Delta wave based on known speciation (@Deltatime),
occurred `r casewavedef$begindate[4]` to
`r as.Date(casewavedef$enddate[4])`. This was eventually followed by the
extremely transmissible Omicron variant, which characterized the wave
lasting `r casewavedef$begindate[6]` to
`r as.Date(casewavedef$enddate[6])`. Finally, while there continued to
be local minima and maxima that could be considered to delineate further
waves, it was decided that the low number of deaths along with
moderation in number of hospitalizations, and regional segmentation in
terms of dispersion around nationwide wave end dates (see Figure 2),
collectively characterized a "post-acute" phase of the pandemic,
starting at `r as.Date(casewavedef$enddate[7])` and extending through
the end of data collection, December 1, 2024. This was driven by
vaccination efforts and milder Omicron subvariants prevalent at the time
and to this day. (@omicron). \newline

![Dispersion of regional minima and maxima about national wave end dates
(cases)](%22../figs/Figure2.png%22){fig-align="center"}

**State performance**

U.S. states exhibited wide variation in case, hospitalization, and death
rates relative to each other in each wave, with the relative performance
changing from earlier to later waves as regional public health responses
diverged.

For instance, New England states and mid-Atlantic states experienced the
greatest burden in terms of deaths and hospitalizations during the first
wave of the pandemic, being highly urban, older, and globally connected
than many other areas of the country. The states of the South Central
and Southeast, on the other hand, had the worst mortality in the second
as well as latest waves (see Figure 3 below). ![State-level mortality by
wave, regionally grouped](%22../figs/Figure3.png%22){fig-align="center"}

The top five states in terms of deaths per population were, in order,
NJ, CT, MA, DC, RI. Seasonal change--i.e., people moving inside at
greater rates in hotter months--brought higher mortality to southern and
southwestern states in the subsequent wave of 2020, with MS, AZ, FL, SC,
and TX occupying the top five spots through September of that year. The
next wave that coincided with the winter of 2020/21 saw SD, ND, OK, AR,
and NM experience the worst state death rates, suggesting that
state-level pandemic response was beginning to play a role in outcomes.
By the final phase of the pandemic (the Omicron subvariants that began
to predominate in spring 2022) this pattern was shown to persist, with
states such as Kentucky and West Virginia consistently occupying the top
five rankings of COVID-19-related mortality while states initially
affected the most--many in New England and the mid-Atlantic--no longer
appearing.

```{r fig.align='center', echo=FALSE, warning=FALSE}

#, fig.subcap=''

kable(statedeathtab, caption = "State-level COVID-19 mortality rank (top 10), by wave start date")

```

**Virulence**

The predominant strain of COVID-19 that characterized each wave became
less virulent, comparing earlier waves to later waves. The original wave
resulted in a mortality : case ratio of
`r round(virulence$deathpercase[1],2)` while the first wave of Omicron
exhibited a ratio of `r round(virulence$deathpercase[6],2)`. By the
approaching-endemicity phase of the pandemic, which we call post-acute,
the ratio had become `r round(virulence$deathpercase[7],2)` when
observing the period March 26, 2022 to October 29, 2022, for which data
reporting is robust. The corresponding hospitalization : case rate
ratios were, respectively, `r round(virulence$deathperhosp[1],2)`,
`r round(virulence$deathperhosp[6],2)`, and
`r round(virulence$deathperhosp[7],2)`. See Figure 4 below for
trendlines of virulence.

![COVID-19 virulence by
wave](%22../figs/Fig4.png%22){fig-align="center"}

# Discussion {#sec-discussion}

This study provides a novel framework for understanding the progression
of the COVID-19 pandemic in the United States by defining waves using a
data-driven approach and evaluating regional and state-level variations
in pandemic dynamics. By identifying seven distinct waves between
January 2020 and December 2024, the findings contribute to a nuanced
understanding of how COVID-19 evolved over time in terms of case rates,
hospitalizations, and virulence.

*Strength of approach*

The definition of pandemic waves using a sliding window approach
highlights the importance of incorporating regional heterogeneity when
evaluating national-level patterns. Unlike prior studies that used fixed
thresholds (e.g., Ayala et al., 2021) or complex statistical models
(e.g., Bali Swain, 2020), this method allowed for a more adaptable
characterization of transitions between waves. The observed alignment of
regional minima and maxima underscores the dynamic relationship between
viral evolution, public health interventions, and sociocultural factors
across the United States.

*Regional stratification*

The analysis of state-level mortality and hospitalization trends further
illustrates significant regional disparities, with urbanized states like
New Jersey and Connecticut experiencing the highest mortality during the
early waves, while southern states such as Mississippi and Kentucky bore
the brunt of later waves. This shift probably is reflective of the
differences in pandemic response measures, and in particular vaccination
rates, or healthcare infrastructure and population behaviors. The
findings also demonstrate a clear decline in virulence over the course
of the pandemic, with mortality-to-case and hospitalization-to-case
ratios decreasing from the initial wave to the post-acute phase
dominated by Omicron subvariants. This trend aligns with previous
research suggesting that increased population immunity and the reduced
severity of later variants contributed to a milder disease profile (Baud
et al., 2020; CDC, 2022).

*Implications of findings*

These results have critical implications for public health planning and
pandemic preparedness. The adaptability of the wave-detection framework
offers a potential tool for monitoring future pandemics, particularly in
scenarios where regional variations and evolving virus characteristics
must be accounted for. Additionally, the state-level disparities in
outcomes emphasize the need for targeted interventions to address
vulnerabilities in specific regions, especially during the early phases
of a pandemic when resources are limited.

*Limitations of the approach*

Several limitations of this study warrant discussion. First, the
reliance on CDC-reported data may introduce biases related to
under-reporting or inconsistencies in data collection across states and
time periods. Second, the study's wave-detection algorithm, while
robust, is inherently sensitive to the choice of parameters (namely,
window size and local minima rather than other indicators of change or
trend) and may not capture subtle transitions between waves. Third, this
analysis does not directly incorporate external factors such as policy
changes, socioeconomic variables, or mobility data, which could further
contextualize the observed patterns.

*Future directions* *for research*

Future research should aim to refine the wave-detection methodology by
utilizing regularly updated data streams or dashboards (such as the
COVID-19 dashboard published by the CDC) as well as by relying on
machine learning techniques to increase predictive power for wave
initiation, peaking, and transition into further viral regimes.
Additionally, longitudinal studies that can incorporate demographic,
healthcare utilization, and even behavioral data could provide deeper
insights into what drove regional disparities. For instance, preexisting
relationships with primary care providers, masking adherence, and
cultural practices like multigenerational living. Finally, examining the
impact of specific public health interventions and vaccination campaigns
on wave dynamics would offer valuable lessons for managing future
pandemics.

For the reasons above, a robust, standardized definition of pandemic
phases provides a useful framework to generate insights into regional
public health responses and their consequences as well as gives a
starting point for future studies of the progression of COVID-19 in the
U.S. and globally.
